---
title: "Getting Started with spectrQC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with spectrQC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# spectrQC: Spectral Unmixing Quality Control

This tutorial walks you through the complete spectrQC workflow for validating spectral flow cytometry unmixing.

```{r}
library(spectrQC)

# Minimal runnable demo used for vignette checks
M_demo <- diag(3)
rownames(M_demo) <- c("FITC", "PE", "APC")
colnames(M_demo) <- c("B2-A", "YG1-A", "R1-A")
W_demo <- derive_unmixing_matrix(M_demo)
stopifnot(all(dim(W_demo) == dim(M_demo)))
```

## Installation


    devtools::install_github("pkheisig/spectrQC")
    library(spectrQC)


## Project Setup

Organize your data with this folder structure:

```
my_project/
├── scc/                          # Single-color control FCS files
│   ├── FITC_beads.fcs
│   ├── PE_beads.fcs
│   └── Unstained.fcs             # Autofluorescence control
├── samples/                      # Experimental FCS files
│   └── Sample1.fcs
└── fcs_control_file.csv          # Control file (optional)
```

## Control File

The control file maps FCS filenames to fluorophore names and peak channels. Generate it using [AutoSpectral](https://github.com/DrCytometer/AutoSpectral):


    # Install AutoSpectral
    if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
    BiocManager::install(c("flowWorkspace", "flowCore", "PeacoQC"))
    devtools::install_github("DrCytometer/AutoSpectral")
    
    # Generate control file
    library(AutoSpectral)
    asp <- get.autospectral.param()
    create.control.file("scc", asp)


Review and edit the generated `fcs_control_file.csv` to ensure correct fluorophore names.

**Manual format:**

| Column | Description |
|--------|-------------|
| `filename` | FCS filename |
| `fluorophore` | Fluorophore name (becomes row name in matrix) |
| `channel` | Peak detector for histogram gating |
| `universal.negative` | TRUE to use as background reference |

***

# Step 1: Build Reference Matrix

Extract spectral signatures from single-color controls:


    library(spectrQC)
    
    # Load control file (optional but recommended)
    control_df <- read.csv("fcs_control_file.csv", stringsAsFactors = FALSE, check.names = FALSE)
    
    # Build reference matrix
    M <- build_reference_matrix(
      input_folder = "scc",
      output_folder = "gating_plots",
      control_df = control_df,
      default_sample_type = "beads"
    )


This creates gating and spectrum plots in `gating_plots/` and exports `reference_matrix.csv`.

### Adjusting Gating Parameters

If auto-gating fails, fine-tune these settings:


    M <- build_reference_matrix(
      input_folder = "scc",
      output_folder = "gating_plots",
      histogram_pct_beads = 0.98,         # Gate width for positive beads
      histogram_pct_cells = 0.35,         # Gate width for cells
      max_clusters = 6,                   # Maximum GMM populations
      gate_contour_beads = 0.999999999,   # Tight bead gate contour
      gate_contour_cells = 0.95,          # Looser cell gate contour
      bead_gate_scale = 1.3               # Ellipse scale factor
    )


***

# Step 2: Generate SCC Report

Review the extracted spectral signatures:


    generate_scc_report(
      M = M,
      scc_dir = "scc",
      output_file = "SCC_QC_Report.pdf"
    )


**Check for:**
- Smooth spectral curves with distinct peaks
- Absence of jagged lines or unexpected peaks
- Low values in the Spectral Spread Matrix

***

# Step 3: Refine the Matrix

Remove debris and outliers using RRMSE-based filtering:


    refined <- refine_scc_matrix(
      M = M,
      scc_dir = "scc",
      rrmse_threshold = 0.05,
      output_dir = "scc_unmixed"
    )
    
    M_final <- refined$M
    W_final <- refined$W


This exports `refined_reference_matrix.csv` and `refined_unmixing_matrix.csv`.

***

# Step 4: Unmix Experimental Samples

Apply the refined matrix to your experimental data:


    unmixed <- unmix_samples(
      sample_dir = "samples",
      M = M_final,
      method = "WLS",
      output_dir = "spectrQC_outputs/unmix_samples"
    )


**Unmixing methods:**

| Method | Description |
|--------|-------------|
| OLS | Ordinary least squares — fast, good for most panels |
| WLS | Weighted least squares — best accuracy, accounts for photon noise |
| NNLS | Non-negative least squares — forces positive abundances |

***

# Interactive Matrix Adjustment

For manual fine-tuning, use the web interface.

## First-time setup (once)

No terminal setup is required for end users. The GUI is bundled with the package.


## Launch the GUI


    launch_gui(
      matrix_dir = getwd(),
      samples_dir = "samples"
    )


This starts both the backend API and bundled frontend automatically, opening `http://localhost:8000`.

For frontend development only:


    launch_gui(
      matrix_dir = getwd(),
      samples_dir = "samples",
      dev_mode = TRUE
    )

**Features:**

- Drag on residual plots to adjust crosstalk coefficients
- Real-time unmixing preview
- Export adjusted matrices

***

# Function Reference

| Function | Purpose |
|----------|---------|
| `build_reference_matrix()` | Extract signatures from single-color controls |
| `generate_scc_report()` | Generate SCC quality report (PDF) |
| `refine_scc_matrix()` | Clean matrix using RRMSE filtering |
| `unmix_samples()` | Unmix experimental FCS files |
| `launch_gui()` | Launch interactive web interface |

***

**Author**: Paul Heisig  
**Contact**: pkheisig@gmail.com

```{r}
sessionInfo()
```
